Informing forest carbon inventories under the Paris Agreement using the Global Forest Carbon Database (ForC v4.0)

### 2024-01-29 doc/manuscript/ForC-EFDB manuscript.Rmd
worked on abstract & intro

<!-- (1) the research conducted or aims, including the rationale-->
- *(Humans have been influencing Earth's climate via interactions with forests for thousands of years, and now)* Forests are recognized as critical to climate change mitigation under the Paris Agreement. 
<!--Forests are recognized as a critical source of additional climate change mitigation needed to meet the goals of the Paris Agreement. -->
However, the efficacy of climate change mitigation planning and reporting depends on having high quality data on forest carbon (C) stocks and changes.
The Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC) is intended to be a definitive source for such data, but needs comprehensive and well-documented data to be so.
<!-- (2) methods-->
- The Global Forest C database (ForC) is a compilation of tens of thousands of ground-based forest C estimates that is fully traceable and publicly available.
Here, we develop and document a process for semi-automated submission of data from ForC into EFDB, <!-- to help ensure that the EFDB has access to the best available data. --> assess the data currently available through ForC, and provide recommendations for improving forest data collection, analysis, and reporting.
<!-- (3) key results-->
- As of `r format(Sys.time(), '%B %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_sent_to_EFDB` of which had undergone necessary review and been submitted.
Records were unevenly distributed across variables (skewed towards aboveground biomass stocks) and geographic regions (skewed towards temperate forests). 
However, `r round((n_records_reviewed_or_added-n_records_sent_to_EFDB)/n_records_sent_to_EFDB,2)*100`% of ForC records reviewed could not be submitted because the original publication lacked information required by the IPCC.
<!-- (4) the main conclusion, including the key points of discussion-->
- In the future, ground-based forest C estimates should be target gaps in the record, and studies should ensure that they report all information necessary for inclusion in EFDB. 
Given that climate change is rapidly impacting the world’s forests, timely reporting of recent estimates will be especially critical to accurate forest C inventories.

@@

Humans have been influencing Earth's climate via ecologically transformative land use practices for >12,000 years [@ellis_people_2021;@sanderman_soil_2017; @bonan_forests_2016].
In recent decades, as anthropogenic climate change has accelerated, this relationship has come into increasing focus [@ipcc_climate_2019;@ipcc_climate_2022].
Deforestation and forest degradation are substantial sources of the greenhouse gas carbon dioxide (CO~2~), currently accounting for >10% of anthropogenic emissions [@friedlingstein_global_2022].
Yet at the same time, CO~2~ uptake by remaining forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 - `r round(5.8*12/44,1)` Gt C yr^-1^ [@xu_changes_2021;@harris_global_2021],
<!--@xu_changes_2021 includes all biomes with trees, estimates ~0.88 Gt C yr^-1^-->
<!--1.6 is net forest estimate in @harris_global_2021 Table 2, which includes only forests-->
which has offset an estimated `r round(.88*44/12/32,2)*100`%<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al. Table 2)--> to `r round(5.8/32,2)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al. Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021].
The future of this important CO~2~ sink depends primarily upon future trajectories of direct human impacts on forests [i.e., deforestation/ degradation vs. conservation/ restoration, [@ipcc_climate_2019;@ipcc_climate_2022], and also upon forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020].

Given this important role of forests in climate regulation, they play a substantial role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015].
Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options [@roe_landbased_2021], with conservation and reforestation having the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].
Yet, envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions through on-the-ground implementation [e.g., @badgley_systematic_2022].
One critical need for ensuring that forest-based climate change mitigation initiatives are effective is realistic planning and reporting, underlain by solid scientific data [@anderson-teixeira_effective_2022; @deng_comparing_2021].

The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change (UNFCCC) [@ipcc_2006_2006;@ipcc_2019_2019].
Under this guidance, greenhouse gas inventories include all managed land, including most of the world's forest land [@ogle_delineating_2018].
The IPCC inventory guidelines include specific instructions for inventories for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [@ipcc_agriculture_2006; @ipcc_2019_2019].
A tiered approach is employed, where the lowest tier (Tier 1) represents the simplest approach and relies on default parameter values -- for example, forest carbon (C) stocks values by ecozone [@fao_global_2012] and forest age class derived as the average of published estimates [@ipcc_2019_2019;@rozendaal_aboveground_2022].
Tier 1 values have improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for improvement as datasets grow and become more widely accessible.
For example, following the 2019 release of the latest IPCC guidelines, it was revealed that IPCC's Tier 1 default failed to capture eight-fold variation of C accumulation in regrowth forests within ecozones [@cook-patton_mapping_2020] and that C stocks in mature African tropical montane forests were two-thirds higher than the IPCC Tier 1 values for these forests [@cuni-sanchez_high_2021].
This rapid evolution of scientific information on C cycling in forests is valuable for informing climate change mitigation efforts but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning. 
Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @cuni-sanchez_high_2021] implies that it is useful for practitioners to have access to locally-specific information, when available.

@@

The goal of this publication is to lower the activation energy required for forest scientists to make their estimates of forest C stocks and changes therein useful and accessible for accounting under IPCC inventory guidelines.
To accomplish this, we document the process of submitting data to EFDB from the Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), which is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021; @anderson-teixeira_forcdb_2023], updated to facilitate data submission to EFDB (ForC v4.0).
We (1) map common scientific forest C estimation methods and definitions to those used by the IPCC;
(2) develop a semi-automated process for preparing ForC data for submission to EFDB;
and (3) assess the data in ForC relevant to EFDB and records that have been submitted to date;
We conclude with recommendations as to how the scientific community can better provide useful data for forest C inventories under the Paris Agreement.

@@

The Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021; @anderson-teixeira_forcdb_2023].
ForC includes data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], and the global soil respiration database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained `r n_records_ForC` records from `r n_plots_ForC` plots in `r n_areas_ForC` distinct geographical areas, along with records of stand age and disturbance history.
As such, ForC is positioned to improve forest-related estimates of CO~2~ emissions and removals through the submission of data to the EFDB.

>>>>

<!-- (1) the research conducted or aims, including the rationale-->
- Humans have been influencing Earth's climate via transformative impacts on forests for >12,000 years, and forests are now recognized as critical to climate change mitigation under the Paris Agreement.
However, the efficacy of climate change mitigation planning and reporting depends on having quality data on forest carbon (C) stocks and changes.
The Emission Factor Database (EFDB) of the International Panel on Climate Change (IPCC) is intended to be a definitive source for such data, but needs comprehensive and well-documented data to be so.
<!-- (2) methods-->
- To facilitate submission of forest C estimates from scientific studies to EFDB, we develop and document a process for semi-automated data submission from the Global Forest C database (ForC v4.0), which is the largest compilation of ground-based forest C estimates. 
We then assess the data currently available through ForC and provide recommendations for improving forest data collection, analysis, and reporting.
<!-- (3) key results-->
- As of `r format(Sys.time(), '%B %Y')`, ForC contained ~`r n_ind_records_EFDB_variables` independent records relevant to EFDB, `r n_records_posted_EFDB` of which had been submitted and posted to EFDB.
Records were unevenly distributed across variables (skewed towards aboveground biomass stocks) and geographic regions (skewed towards temperate forests).
`r round((n_records_reviewed_or_added-n_records_sent_to_EFDB)/n_records_sent_to_EFDB,2)*100`% of ForC records reviewed could not be submitted because the original publication lacked required information.
<!-- (4) the main conclusion, including the key points of discussion-->
- In the future, ground-based forest C estimates should target gaps in the record, and studies should ensure that they report all information necessary for inclusion in EFDB. 
Given that climate change is rapidly impacting the world’s forests, timely reporting of recent estimates will be especially critical to accurate forest C inventories.

@@

Humans have been influencing Earth's climate via ecologically transformative land use practices for >12,000 years [@ellis_people_2021;@sanderman_soil_2017; @bonan_forests_2016].
In recent decades, as both anthropogenic land transformation and climate change have accelerated, this relationship has come into increasing focus [@ipcc_climate_2019].
Deforestation and forest degradation are substantial sources of the greenhouse gas carbon dioxide (CO~2~), currently accounting for >10% of anthropogenic emissions [@friedlingstein_global_2022].
Yet at the same time, CO~2~ uptake by remaining forests, woodlands, and savannas has exceeded releases from deforestation and other severe disturbances, resulting in a net carbon CO~2~ sink of ~0.88 - `r round(5.8*12/44,1)` Gt C yr^-1^ [@xu_changes_2021;@harris_global_2021],
<!--@xu_changes_2021 includes all biomes with trees, estimates ~0.88 Gt C yr^-1^
-1.6 is net forest estimate in @harris_global_2021 Table 2, which includes only forests-->
ofsetting an estimated `r round(.88*44/12/32,2)*100`<!--net forest estimate from xu_changes_2021 / fossil fuel & cement, (Harris et al. Table 2)--> - `r round(5.8/32,2)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al. Table 2)--> of anthropogenic CO~2~ emissions from fossil fuels and cement [@xu_changes_2021;@harris_global_2021].
The future of this important CO~2~ sink depends primarily upon future trajectories of direct human impacts on forests [i.e., deforestation/ degradation vs. conservation/ restoration, @ipcc_climate_2019;@ipcc_climate_2022], and also upon forest responses to climate change, which are likely to reduce the sink strength [@mcdowell_pervasive_2020].

In recognition of their important role in climate regulation, forests play a substantial role in international plans for climate change mitigation under the Paris Agreement [@unfccc_adoption_2015; @grassi_key_2017].<!--As of 2016, forest-based mitigation accounted for 26% of total planned greenhouse gas mitigation within Nationally Determined Contributions under the Paris Agreement [@grassi_key_2017].-->
Forest conservation, reforestation, and improved sustainable management all have significant -- and relatively cost-effective -- potential as climate change mitigation options [@roe_landbased_2021], with conservation and reforestation having the fourth and fifth largest net emission reduction potentials or all mitigation options [@ipcc_summary_2022].
Yet, envisioned forest-based climate change mitigation initiatives do not always correspond to actual emission reductions through on-the-ground implementation [e.g., @badgley_systematic_2022].
One critical need for ensuring that forest-based climate change mitigation initiatives are effective is realistic planning and reporting, underlain by solid scientific data [@anderson-teixeira_effective_2022; @deng_comparing_2021].

The International Panel on Climate Change (IPCC) provides guidance for national greenhouse gas inventories for reporting to the United Nations Framework Convention on Climate Change (UNFCCC) [@ipcc_2006_2006;@ipcc_2019_2019].
Under this guidance, greenhouse gas inventories include all managed land, which includes most of the world's forest land [@ogle_delineating_2018; @ellis_anthropogenic_2010].
The IPCC inventory guidelines include specific instructions for inventories for greenhouse gas (mainly CO~2~) exchanges between forest land and the atmosphere [@ipcc_agriculture_2006; @ipcc_2019_2019].
A tiered approach is employed, where the lowest tier (Tier 1) represents the simplest approach and relies on default parameter values -- for example, forest carbon (C) stocks values by ecozone [@fao_global_2012] and forest age class derived as the average of published estimates [@ipcc_2019_2019;@rozendaal_aboveground_2022].
Tier 1 values have improved over the years as more of the relevant underlying data has become available [@requenasuarez_estimating_2019; @rozendaal_aboveground_2022], but there remains room for improvement as datasets grow and become more widely accessible.
For example, following the 2019 release of the latest IPCC guidelines, it was revealed that IPCC's Tier 1 default failed to capture eight-fold variation of C accumulation in regrowth forests within ecozones [@cook-patton_mapping_2020] and that C stocks in mature African tropical montane forests were two-thirds higher than the IPCC Tier 1 values for these forests [@cuni-sanchez_high_2021].
This rapid evolution of scientific information on C cycling in forests is valuable for informing climate change mitigation efforts but requires improved mechanisms for communicating the latest information from scientific researchers to the practitioners who need reliable estimates for greenhouse gas mitigation planning. 
Moreover, high variability of forest C cycling within ecozones [e.g., @cook-patton_mapping_2020; @cuni-sanchez_high_2021] implies that it is useful for practitioners to have access to locally-specific information, when available.

@@

The goal of this publication is to facilitate submission of forest C estimates from scientific studies to EFDB.
We document the process of submitting data to EFDB from the Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), which is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021; @anderson-teixeira_forcdb_2023].
We (1) map common scientific forest C estimation methods and definitions to those used by the IPCC;
(2) develop a semi-automated process for preparing ForC data for submission to EFDB;
and (3) assess the data in ForC relevant to EFDB and records that have been submitted to date;
We conclude with recommendations as to how the scientific community can better provide useful data for forest C inventories under the Paris Agreement.

@@

The Global Forest Carbon Database, ForC ([https://forc-db.github.io/](https://forc-db.github.io/)), is the largest collection of published estimates of forest C stocks, increments, and annual fluxes [@anderson-teixeira_forc_2018; @anderson-teixeira_carbon_2021; @anderson-teixeira_forcdb_2023].
ForC includes data ingested from individual publications and relevant databases, including the Global Reforestation Opportunity Assessment (GROA) database [@cook-patton_mapping_2020, database doi: 10.5281/zenodo.3983644], and the global soil respiration database [SRDB-V5, @bond-lamberty_global_2010; @jian_restructured_2021].
As of `r format(Sys.time(), '%B %d, %Y')`, ForC contained `r n_records_ForC` records from `r n_plots_ForC` plots in `r n_areas_ForC` distinct geographical areas, along with records of stand age and disturbance history.
As such, ForC is positioned to improve forest-related estimates of CO~2~ emissions and removals through the submission of data to the EFDB.
<!--ForC is fully traceable and publicly available.-->
